---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# lite

[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/paulnorthrop/lite?branch=main&svg=true)](https://ci.appveyor.com/project/paulnorthrop/lite)
[![R-CMD-check](https://github.com/paulnorthrop/lite/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/paulnorthrop/lite/actions/workflows/R-CMD-check.yaml)
[![Coverage Status](https://codecov.io/github/paulnorthrop/lite/coverage.svg?branch=main)](https://codecov.io/github/paulnorthrop/lite?branch=main)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/lite)](https://cran.r-project.org/package=lite)
[![Downloads (monthly)](https://cranlogs.r-pkg.org/badges/lite?color=brightgreen)](https://cran.r-project.org/package=lite)
[![Downloads (total)](https://cranlogs.r-pkg.org/badges/grand-total/lite?color=brightgreen)](https://cran.r-project.org/package=lite)

## Likelihood-Based Inference for Time Series Extremes

The **lite** package performs likelihood-based inference for stationary time series extremes.  The general approach follows [Fawcett and Walshaw (2012)](https://doi.org/10.1002/env.2133). There are 3 independent parts to the inference, all performed using maximum likelihood estimation.

1. A Bernoulli ($p_u$) model for whether a given observation exceeds the threshold $u$.
2. A generalised Pareto, GP ($\sigma_u, \xi$), model for the marginal distribution of threshold excesses.
3. The $K$-gaps model for the extremal index $\theta$, based on inter-exceedance times.

For parts 1 and 2 it is necessary to adjust the inferences because we expect that the data will exhibit cluster dependence.  This is achieved using the methodology developed in [Chandler and Bate (2007)](https://doi.org/10.1093/biomet/asm015) to produce a log-likelihood that is adjusted for this dependence.  This is achieved using the [chandwich package](https://cran.r-project.org/package=chandwich). For part 3, the methodology described in [Süveges and Davison (2010)](https://doi.org/10.1214/09-AOAS292) is used, implemented by the function `kgaps` in the [exdex package](https://cran.r-project.org/package=exdex).  The (adjusted) log-likelihoods from parts 1, 2 and 3 are combined to make inferences about return levels.

### An example: Cheeseboro wind gusts

The function `flite` makes inferences about $(p_u, \sigma_u, \xi, \theta)$.  We illustrate this using the `cheeseboro` data from the [exdex package](https://cran.r-project.org/package=exdex), which contains hourly wind gust data from each January over the 10-year period 2000-2009.

```{r, echo = FALSE}
got_exdex <- requireNamespace("exdex", quietly = TRUE)
```

First, we make inferences about the model parameters.

```{r model, eval = got_exdex}
library(lite)
cdata <- exdex::cheeseboro
# Each column of the matrix cdata corresponds to data from a different year
# flite() sets cluster automatically to correspond to column (year)
cfit <- flite(cdata, u = 45, k = 3)
```

Then, we make inferences about the 100-year return level, including 95\% confidence intervals.  The argument `ny` sets the number of observations per year, which is $31 \times 24 = 744$ for these data.

```{r returnLevels, eval = got_exdex}
rl <- returnLevel(cfit, m = 100, level = 0.95, ny = 31 * 24)
rl
```

### Installation

To get the current released version from CRAN:

```{r installation, eval = FALSE}
install.packages("lite")
```

### Vignette

See `vignette("introduction-to-lite", package = "lite")` for an overview of the package.
